N-dimensional Tic Tac Toe, and Adventure in Modules

### 2017-05-12 paper/paper.tex
coupling section

    end
  \end{verbatim}
  We decided against this approach i.e. isolating IO into a separate module,
  because it began to pollute our design's dependency graph with unwanted edges.
  In haskell, type class instances exist in an global overloaded name space,
  that takes advantage of haskell's dispatch system. So each data type that is
  implemented for the \texttt{Show} typeclass is able to be printed to StdOut
  \textit{without} carrying around its respective show function. This is not the
  case in SML, and it was sorely missed in our implementation. If we were to
  follow the Haskell style of IO (separating impure and pure code explicitly),
  then an IO dependency would be requisite anytime we needed to perform any IO.
  Thus, the IO node would become a dominating node in our dependency graph, and
  the its import into every functor would quickly become boilerplate and
  unweidly. Hence, we chose an SML style design, where the structures that are
  mapped upon carry any relevant functions with them, which maintains a clean
  decomposition in SML. This type of structure bloat, although disgraceful to a
  haskell programming, seems to be encouraged by SML's module system.

  \subsection{You can do it in 2-dimensions, but can you do it in n-dimensions!}
  Our initial intent was to implement a pokemon battle simulator. As we
  dove into the problem domain we realized that instead of falling into a
  specific, flow-chart design with modules like: Pokemon, Combat, Moves etc. it
  would be better design to create an abstract game engine and then make pokemon
  \textit{an instance of} the abstract game engine. A similar pattern occurred
  during our first implementation of the matrix library utilized by the tic tac
  toe state. We wrote a single matrix module with more rigidly defined 
  dimensions and underlying implementation details. While it was abstract, it was not very modular,
  as it required writing a new module from the ground up if any of the 
  implementation details were to be changed. Thus, in the spirit of modularity, we refactored the matrix
  library to implement an abstract \texttt{CONTAINER} signature. Then based on
  this signature, it became possible to create sub-signatures that specified the
  types of the CONTAINER signature. A good example is shown below:

@@

  \end{verbatim}
  In this example we show the relevant parts of the \texttt{CONTAINER} signature
  and how we used that signature to create a \textit{more specific}
  \texttt{SQUAREMATRIX} signature e.g. a matrix whose size datatype need only be
  a single integer value. This allows for a very flexible and modular
  implementation, in addition to square matrices we could create jagged two-dimensional matrices
  by following the same pattern viz. defining a \texttt{JAGGEDMATRIX} and
  setting size to ``int * int''.

  Not only is the second layer of abstraction useful for changing matrix
  properties on the fly, but it is also useful for changing the underlying
  implementation and dimensionality of the matrix. With this design we were able
  to create a functor that, when given a structure of type
  \texttt{SQUAREMATRIX}, and a structure of type \texttt{VECT} will return a
  matrix of whose underlying datatype is implemented by a structure, V of type
  \texttt{VECT}, that dictates the underlying datatype of the matrix, and a
  structure M, of type \texttt{SQUAREMATRIX} that specifies the datatype that
  stores the V structure. For example, we create matrices that are created with
  arrays, vectors and lists by passing each as a structure of type
  \texttt{VECT}; any such data structure is viable as long as it supports the
  \texttt{VECT} signature. By altering the structure M, of type
  \texttt{SQUAREMATRIX} we create n-dimensional matrices. So by varying both we
  gain the ability to create n-dimensional matrices, with abstracted underlying
  data types. Thus, the double abstracted design of our matrix library is a
  semantic definition of a matrix, which separates and abstracts the underlying
  datatype implementation from the dimensionality of the matrix and the
  underlying implementation details, thereby allowing for extreme reuse and
  flexibility.

  \subsection{The Functor is love, the Functor is life}
  % How functors enabled our clean decomposition and modular design

>>>>

    end
  \end{verbatim}
  We decided against this approach i.e. isolating IO into a separate module,
  because it began to pollute our design's dependency graph with unwanted edges.
  In Haskell, type class instances exist in an global overloaded name space,
  that takes advantage of Haskell's dispatch system. So each data type that is
  implemented for the \texttt{Show} typeclass is able to be printed to StdOut
  \textit{without} carrying around its respective show function. This is not the
  case in SML, and it was sorely missed in our implementation. If we were to
  follow the Haskell style of IO (separating impure and pure code explicitly),
  then an IO dependency would be requisite anytime we needed to perform any IO.
  Thus, the IO node would become a dominating node in our dependency graph, and
  the its import into every functor would quickly become boilerplate and
  unwieldy. Hence, we chose an SML style design, where the structures that are
  mapped upon carry any relevant functions with them, which maintains a clean
  decomposition in SML. This type of structure bloat, although disgraceful to a
  Haskell programming, seems to be encouraged by SML's module system.

  \subsection{You can do it in 2-dimensions, but can you do it in n-dimensions!}
  Our initial intent was to implement a pokemon battle simulator. As we
  dove into the problem domain we realized that instead of falling into a
  specific, flow-chart design with modules like: Pokemon, Combat, Moves etc. it
  would be better design to create an abstract game engine and then make pokemon
  \textit{an instance of} the abstract game engine. A similar pattern occurred
  during our first implementation of the matrix library utilized by the tic tac
  toe state. We wrote a single matrix module with more rigidly defined 
  dimensions and underlying implementation details. While it was abstract, it was not very modular,
  as it required writing a new module from the ground up if any of the 
  implementation details were to be changed. Thus, in the spirit of modularity, we refactored the matrix
  library to implement a \texttt{CONTAINER} signature that specifies the interface for
  a container abstract data type. As above, we were able to use structural matching and the
  \texttt{include} keyword to extend our container ADT to more specific kinds of container. A good
  example is shown below:

@@

  \end{verbatim}
  In this example we show the relevant parts of the \texttt{CONTAINER} signature
  and how we used that signature to create a \textit{more specific}
  \texttt{SQUAREMATRIX} signature e.g. a matrix whose size datatype need only be
  a single integer value. Similarly, we can define a \texttt{VECT} signature that describes
  containers that can be used as one-dimensional vectors (like list or arrays) by restricting 
  both the \texttt{size} and \texttt{index} types to be \texttt{int}.
  This allows for a very flexible and modular
  implementation, in addition to square matrices we could create jagged two-dimensional matrices
  by following the same pattern viz. defining a \texttt{JAGGEDMATRIX} and
  setting size to ``int * int''.

  One consequence of this design is that you can create n-dimensional square matrices given only
  modules implementing \texttt{VECT} and a functor. We define a functor \texttt{SquareMatrixFn}
  that takes a module \texttt{M} implementing \texttt{SQUAREMATRIX} and a module \texttt{V}
  implementing \texttt{VECT} and produces a new square matrix of dimension n+1 where n is the
  dimensionality of M. We simply use \texttt{V} to store the polymorphic matrix elements and \texttt{M} to
  store rows of \texttt{V}. The type of indices of our new matrix is just the product of the index types
  of its components.

  Because all \texttt{VECT} modules also implement \texttt{SQUAREMATRIX}, we
  can create two-dimensional matrices by passing any two \texttt{VECT} implementations to our functor.
  From there, we can create n-dimensional matrices by simply passing the result back into the functor
  as \texttt{M} any number of times. Therefore, given only a module implementing \texttt{VECT}, our
  functor can automatically produce modules representing matrices of n-dimensions.
  Because functors are not first-class, we cannot automate this process by passing in
  the intended dimensionality as an integer argument, so instead we are stuck threading the results
  of our functor back into itself. This could prove tedious for matrices with high dimensionality, but for our
  purposes was acceptable.

  This use of functors also allows for mixing of containers in matrices.
  We implemented \texttt{VECT} for the basis array, list, and vector libraries. Our functor can take any
  combination of these and produce a matrix module that works the same
  as any other. As these implementation details are hidden by the container ADT, we can safely mix
  any of them to form another valid container matrix.

  All of this has the result that our \textit{Tic-tac-toe} implementation works for matrices of any
  dimensionality and implemented using a variety of data structures.

  \subsection{Coupling}

  One area where our implementation encountered some issues was in the coupling of our system.
  Despite our best efforts, our modules are not as loosely coupled as we would like. Each module
  carries with it the modules that it depends on as part of its structure. This is because many of the
  interfaces depend on type definitions contained in the modules that they depend on (for example,
  \texttt{ACTION} modules depend on the definition of the \texttt{state} type defined in \texttt{STATE}).
  This is not ideal in terms of
  coupling, and forced us to use a lot of trivial functors that thread these dependencies through our
  system.

  Resolving this is a tricky problem. If we make the types of our abstract game interface all concrete,
  we could gain looser coupling at the cost of how generic our abstract game engine can be. Concrete
  type definitions in all signatures would alleviate this problem, but we would wind up unable to extend
  our abstract interface to any other games aside from \textit{Tic-tac-toe}. Being locked into a single
  implementation in this way would simply reduce our modules to simple namespaces with none of the 
  abstraction or extensibility benefits associated with modular programming.

  One possible solution to this problem could be to somehow serialize interactions between
  different modules into a standard format (much like how web servers and clients often convert data
  into JSON to pass back and forth). However, this still seems to include some notion of coupling, just
  deferred now to serialization and deserialization processes. A module would still have to know how to
  serialize and deserialize information and how to interpret the communications it receives from 
  other modules. So while we would decouple at the type level, we would still retain coupling at the
  informational level. It is unclear if there is a cleaner decomposition of our game domain that would
  allow for looser coupling while still retaining similar abstraction and cohesion as our current system.
